---
title: "Annotating nextflow processed data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{annotating-nextflow-processed-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Intro

This vignette documents in-practice usage of the annotation utils for nf-processed data files;
it will probably updated as practices or underlying functions (which are still experimental) change. 

Since examples reference real data, most of the interactive portions need at least read access.
The portions that require modify access are mostly illustrative; `eval=F` by default. 
(Note that if these are rerun, annotations will be duplicated.) 

## Set up

First load the `nfportalutils` package and log in. 
The recommended default usage of `syn_login` is to use it without directly passing in credentials. 
Instead, have available the `SYNAPSE_AUTH_TOKEN` environment variable with your token stored therein. 
```{r setup}
library(nfportalutils)
library(data.table)
syn_login()
```

## nf-rnaseq 

### Creating helper resource files

To add annotations and provenance, we need a reference mapping of inputs and outputs involved in the workflow.
The below puts together a table representing this mapping by parsing the samplesheet in combination with the output directory. 
For the nextflow rna-seq workflow, the samplesheet is an output file in "pipeline_info".
```{r rnaseq-resources, eval=FALSE}
sample_io <- map_sample_io(workflow = "nf-rnaseq",
                           samplesheet = "syn30841441", # synapse file
                           syn_out = "syn30840584")
```


### Add provenance

Add provenance for the files involved using `add_activity_batch`. 

Note: Why is workflow name already available but not workflow link?
The workflow link is a more specific reference and includes version numbers. 

```{r rnaseq-add-provenance, eval=FALSE}

wf_link <- "https://nf-co.re/rnaseq/3.7/output#star-and-salmon"
prov <- add_activity_batch(sample_io$output_id, 
                           sample_io$workflow, 
                           wf_link,
                           sample_io$input_id)
```


### Add annotations

Annotations expected for aligned reads output are different from quantified expression output. 
This is handled by a couple of specialized functions.

#### Aligned reads data files

First, it helps to see the template for the aligned reads`(.bam`) data.  
```{r rnaseq-see-annotations-bam}

template <- "bts:ProcessedAlignedReadsTemplate"
schema <- "https://raw.githubusercontent.com/nf-osi/nf-metadata-dictionary/main/NF.jsonld"
props <- get_dependency_from_json_schema(id = template)
props
```

This looks like a large number of different annotations to put together (though most are optional). 
Fortunately, the utility `annotate_aligned_reads` will try to do much of the work. 
It will look for the `.bam` files and generate annotations for those `.bam` files.
The template and schema mentioned above are the defaults so don't have to be specified explicitly. 
It's recommended to read the docs (`?annotate_aligned_reads`) for details on how annotations are compiled; 
usage of `annotate_aligned_reads` does still require knowing what should be filled in manually.

Note: Why is workflow info added as both provenance and annotations? Isn't that redundant?
The answer is that the portal uses only annotations and not provenance/annotations are more queryable currently.

```{r rnaseq-make-annotations-bam, eval=FALSE}

l2_manifest <- annotate_aligned_reads(sample_io,
                                      samtools_stats_file = "syn30841284",
                                      update = FALSE)

l2_manifest[, genomicReference := "GRCh38"]
l2_manifest[, workflowLink := "https://nf-co.re/rnaseq/3.7/output#star-and-salmon"]

```


This looks good, so we can go ahead and apply the annotations:
```{r rnaseq-apply-annotations-l2, eval=FALSE}

annotate_with_manifest(l2_manifest)
```

TO DO: Instead of submitting as above, if we prefer to submit via schematic/DCA, we can make this a schematic-compatible manifest.

#### Expression data files

Another convenience function is provided to help annotate gene expression data files.

```{r rnaseq-add-annotations-l3, eval=FALSE}

l3_manifest <- annotate_expression(sample_io, update = FALSE)
l3_manifest[, workflowLink := "https://nf-co.re/rnaseq/3.7/output#star-and-salmon"]
annotate_with_manifest(l3_manifest)
```


## nf-sarek

As the second part of this vignette shows, the process is pretty much the same as for [nf-rnaseq](#nf-rnaseq),
though mapping inputs-outputs can be somewhat slower because of the output structure.

### Creating helper resource files
```{r sarek-resources, eval=FALSE}

sample_io <- map_sample_io(workflow = "nf-sarek",
                           samplesheet = "PilotBatch-WES-samplesheet.txt", # local file
                           syn_out = "syn27650634")
# Modify workflow assignment a bit
sample_io[workflow =="Strelka", workflow := "Strelka2"]
```

### Add provenance

Note: For germline, make sure to include an additional DeepVariant reference as well.
```{r sarek-add-provenance, eval=FALSE}

wf_link <- c(FreeBayes = "https://nf-co.re/sarek/2.7.1/output#freebayes",
             Mutect2 = "https://nf-co.re/sarek/2.7.1/output#gatk-mutect2",
             Strelka2 = "https://nf-co.re/sarek/2.7.1/output#strelka2")
add_activity_batch(sample_io$output_id, 
                   sample_io$workflow, 
                   wf_link[sample_io$workflow], 
                   sample_io$input_id)
   
```

### Add annotations

Adding annotations can be done with `annotate_called_variants`, which can be used for both somatic and germline.
 
```{r sarek-make-annotations, eval=FALSE}

annotations <- annotate_called_variants(sample_io, data_type = "SomaticVariants")
annotations[, workflowLink := wf_link[workflow]]
```

And apply...
```{r sarek-apply-annotations, eval=FALSE}

annotate_with_manifest(annotations)
```



