---
title: "Bringing Portal Data to Other Platforms: cBioPortal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bringing Portal Data to Other Platforms: cBioPortal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**Document Status:** Draft  
**Estimated Reading Time:** 8 min

## Special acknowledgments 

Functionality demonstrated in this vignette benefited greatly from code originally written by [hhunterzinck](https://github.com/hhunterzinck). 

## Intro

This describes how to package some Synapse processed data as a cBioPortal study dataset.
A cBioPortal study contains one or more data types, see [cBioPortal docs](https://docs.cbioportal.org/file-formats/). 
The current API covers creating a cBioPortal study with a subset of data types relevant to the NF workflow (so not all data types). 
The design has been inspired by and should feel somewhat like working with the R package [usethis](https://github.com/r-lib/usethis), 
and data types can be added to the study package interactively.

Though there is some checking depending on the data type, final [validation](#Validation) with the official cBioPortal validation tools/scripts should still be run.

Breaking changes are possible as the API is still in development.

## Set up

First load the `nfportalutils` package and log in. 
The recommended default usage of `syn_login` is to use it without directly passing in credentials. 
Instead, have available the `SYNAPSE_AUTH_TOKEN` environment variable with your token stored therein. 

```{r setup, eval=FALSE}
library(nfportalutils)
syn_login()
```

## Create a new study dataset

First create the study dataset "package" where we can put together the data. 
Each study dataset combines multiple data types -- clinical, gene expression, gene variants, etc.

```{r cbp_new_study, eval=FALSE}

cbp_new_study(cancer_study_identifier = "npst_nfosi_ntap_2022",
              name = "Plexiform Neurofibroma and Neurofibroma (Pratilas 2022)",
              citation = "TBD")
```

## Add data types to study

Data types can be most easily added _in any order_ using the `cbp_add*` functions.
These functions download data files and create the meta for them. 

Note that:

- These should be run with the working directory set to the study dataset directory as set up above to ensure consistent metadata.
- **Defaults are for known NF-OSI processed data outputs**. 
- If these defaults don't apply because of changes in the scenario, take a look at the lower-level utils `make_meta_*` or edit the files manually after.
- Data types can vary in how much additional work is needed in remapping, reformatting, custom sanity checks, etc.

### Add mutations data

- `maf_data` references a final merged maf output file from the NF-OSI processing pipeline OK for public release. 
- This data file type requires no further modifications except renaming.

```{r add_maf, eval=FALSE}

maf_data <- "syn36553188"

add_cbp_maf(maf_data)
```

### Add copy number alterations (CNA) data

- `cna_data` is expected to be a `.seg` file on Synapse.

```{r add_cna, eval=FALSE}

cna_data <- "syn********"

cbp_add_cna(cna_data)
```

### Add expression data

- `expression_data` is expected to be a `.txt` called `gene_tpm.tsv` file on Synapse.
- The NF-OSI default includes including the raw expression data as well, called `gene_counts.tsv`, but this can be omitted.
- These NF-OSI outputs will be somewhat modified in translation to have the required headers.

```{r add_expression, eval=FALSE}

mrna_data <- "syn********"
mrna_data_raw <- "syn********"

cbp_add_expression(mrna_data,
                   expression_data_raw = mrna_data_raw)
```

### Add clinical data

- `clinical_data` is a prepared clinical data table already subsetted to those released in this study, or pass in a query that can be used for subsetting if using a full clinical database table. For example, the full clinical cohort comprises patients 1-50, but this study dataset consists of available and releasable data only for patients 1-20 for expression data and data patients 15-20 for cna data. Here, `clinical_data` can be a smaller table of just those 1-30, or it can be the original table but pass in a suitable additional filter, e.g. `where release = 'batch1'`.
- Clinical data requires mapping to be as consistent with other public datasets as possible. `ref_map` defines the mapping of clinical variables from the NF-OSI data dictionary to cBioPortal's. Only variables in the mapping are exported to cBioPortal. Follow link below to inspect the default file and format used.
- Clinical data should be added last for overall sample checks to work. For example, if there is expression data for patients 1-20 and cna data patients 15-20,
it can more informatively warn about any missing/mismatches. 

```{r add_clinical, eval=FALSE}

clinical_data <- "select * from syn43278088"
ref_map <- "https://raw.githubusercontent.com/nf-osi/nf-metadata-dictionary/main/mappings/cBioPortal.yaml"

cbp_add_clinical(clinical_data, ref_map)
```

## Validation

There are additional steps such as generating case lists and validation that have to be done _outside_ of the package with a cBioPortal backend, where each portal may have specific configurations (such as genomic reference) to validate against.
See the [general docs for dataset validation](https://docs.cbioportal.org/using-the-dataset-validator/).

For the _public_ portal, the suggested step using the public server is given below.  

Assuming your present working directory is `~/datahub/public` and a study folder called `npst_nfosi_ntap_2022` has been placed into it, mount the dataset into the container and run validation with:  
`docker run --rm -v $(pwd):/datahub cbioportal/cbioportal:5.4.7 validateStudies.py -d /datahub -l npst_nfosi_ntap_2022 -u http://cbioportal.org -html /datahub/npst_nfosi_ntap_2022/html_report`

The html report will list issues by data types to help with any corrections needed.  
