---
title: "Package Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{package-installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation {.tabset}

THIS VERSION:
This is the recommended installation for the [current release](https://github.com/nf-osi/nfportalutils/releases/tag/v0.9500-presynapser) of `nfportalutils` **that does not depend on `synapser`**.


There *will* be a **newer version of `nfportalutils` with `synapser`** (i.e. see [this issue](https://github.com/nf-osi/nfportalutils/issues/94) on the roadmap), and instructions will be accordingly revised when the refactor is completed.

Make sure you are in a new R session (in RStudio, go to `Session > Restart R` if necessary).

### Windows

#### Outside R

- Install a supported `python` version (recommend Python 3.11) by downloading installation binary at https://www.python.org/downloads/windows/.
- Install `synapseclient` following instructions at https://pypi.org/project/synapseclient/4.0.0/.

#### With R
- Install `remotes` to help install other packages: 
`install.packages("remotes")`.

- Check that you have the correct `reticulate` version: 
`packageVersion("reticulate")`.  
If the result is anything other than `1.28`, then go to the `reticulate` install step, otherwise skip to `nfportalutils` install.

- Install `reticulate`:
```
> remotes::install_github("rstudio/reticulate@v1.28")
Downloading GitHub repo rstudio/reticulate@v1.28
Running `R CMD build`...
* checking for file 'C:\Users\Erik\AppData\Local\Temp\Rtmporbql0\remotes322031ec7c50\rstudio-reticulate-3de77d1/DESCRIPTION' ... OK
* preparing 'reticulate':
* checking DESCRIPTION meta-information ... OK
* cleaning src
* checking for LF line-endings in source and make files and shell scripts
* checking for empty or unneeded directories
* building 'reticulate_1.28.tar.gz'
Installing package into ‘C:/Users/Erik/AppData/Local/R/win-library/4.3’
(as ‘lib’ is unspecified)
* installing *source* package 'reticulate' ...
** using staged installation
** libs
using C++ compiler: 'G__~1.EXE (GCC) 12.3.0'
g++ -std=gnu++17  -I"C:/PROGRA~1/R/R-43~1.2/include" -DNDEBUG  -I'C:/Users/Erik/AppData/Local/R/win-library/4.3/Rcpp/include'   -I"C:/rtools43/x86_64-w64-mingw32.static.posix/include"     -O2 -Wall  -mfpmath=sse -msse2 -mstackrealign  -c RcppExports.cpp -o RcppExports.o
g++ -std=gnu++17  -I"C:/PROGRA~1/R/R-43~1.2/include" -DNDEBUG  -I'C:/Users/Erik/AppData/Local/R/win-library/4.3/Rcpp/include'   -I"C:/rtools43/x86_64-w64-mingw32.static.posix/include"     -O2 -Wall  -mfpmath=sse -msse2 -mstackrealign  -c event_loop.cpp -o event_loop.o
g++ -std=gnu++17  -I"C:/PROGRA~1/R/R-43~1.2/include" -DNDEBUG  -I'C:/Users/Erik/AppData/Local/R/win-library/4.3/Rcpp/include'   -I"C:/rtools43/x86_64-w64-mingw32.static.posix/include"     -O2 -Wall  -mfpmath=sse -msse2 -mstackrealign  -c libpython.cpp -o libpython.o
g++ -std=gnu++17  -I"C:/PROGRA~1/R/R-43~1.2/include" -DNDEBUG  -I'C:/Users/Erik/AppData/Local/R/win-library/4.3/Rcpp/include'   -I"C:/rtools43/x86_64-w64-mingw32.static.posix/include"     -O2 -Wall  -mfpmath=sse -msse2 -mstackrealign  -c output.cpp -o output.o
g++ -std=gnu++17  -I"C:/PROGRA~1/R/R-43~1.2/include" -DNDEBUG  -I'C:/Users/Erik/AppData/Local/R/win-library/4.3/Rcpp/include'   -I"C:/rtools43/x86_64-w64-mingw32.static.posix/include"     -O2 -Wall  -mfpmath=sse -msse2 -mstackrealign  -c python.cpp -o python.o
g++ -std=gnu++17  -I"C:/PROGRA~1/R/R-43~1.2/include" -DNDEBUG  -I'C:/Users/Erik/AppData/Local/R/win-library/4.3/Rcpp/include'   -I"C:/rtools43/x86_64-w64-mingw32.static.posix/include"     -O2 -Wall  -mfpmath=sse -msse2 -mstackrealign  -c readline.cpp -o readline.o
g++ -std=gnu++17  -I"C:/PROGRA~1/R/R-43~1.2/include" -DNDEBUG  -I'C:/Users/Erik/AppData/Local/R/win-library/4.3/Rcpp/include'   -I"C:/rtools43/x86_64-w64-mingw32.static.posix/include"     -O2 -Wall  -mfpmath=sse -msse2 -mstackrealign  -c signals.cpp -o signals.o
g++ -std=gnu++17 -shared -s -static-libgcc -o reticulate.dll tmp.def RcppExports.o event_loop.o libpython.o output.o python.o readline.o signals.o -LC:/rtools43/x86_64-w64-mingw32.static.posix/lib/x64 -LC:/rtools43/x86_64-w64-mingw32.static.posix/lib -LC:/PROGRA~1/R/R-43~1.2/bin/x64 -lR
installing to C:/Users/Erik/AppData/Local/R/win-library/4.3/00LOCK-reticulate/00new/reticulate/libs/x64
** R
** inst
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
*** copying figures
** building package indices
** installing vignettes
** testing if installed package can be loaded from temporary location
** testing if installed package can be loaded from final location
** testing if installed package keeps a record of temporary installation path
* DONE (reticulate)
```

- Install `nfportalutils` (choose option `3 - None`): 
```
> remotes::install_github("nf-osi/nfportalutils@v0.9500-presynapser")
Downloading GitHub repo nf-osi/nfportalutils@v0.9500-presynapser
These packages have more recent versions available.
It is recommended to update all of them.
Which would you like to update?

1: All                               
2: CRAN packages only                
3: None                              
4: reticulate (1.28 -> 1.35.0) [CRAN]
```

```
Running `R CMD build`...
* checking for file 'C:\Users\Erik\AppData\Local\Temp\RtmpQzT2JT\remotes158c6d2c7833\nf-osi-nfportalutils-61636cd/DESCRIPTION' ... OK
* preparing 'nfportalutils':
* checking DESCRIPTION meta-information ... OK
* checking for LF line-endings in source and make files and shell scripts
* checking for empty or unneeded directories
Omitted 'LazyData' from DESCRIPTION
* building 'nfportalutils_0.9500.tar.gz'
Installing package into ‘C:/Users/Erik/AppData/Local/R/win-library/4.3’
(as ‘lib’ is unspecified)
* installing *source* package 'nfportalutils' ...
** using staged installation
** R
** inst
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
** building package indices
** installing vignettes
** testing if installed package can be loaded from temporary location
** testing if installed package can be loaded from final location
** testing if installed package keeps a record of temporary installation path
* DONE (nfportalutils)
```


#### Post-installation

It is recommended to set your Synapse authentication token as an environment variable that R can access.
You can create an `.Renviron` in your home. For example, create 'C:/Users/Erik/Documents/.Renviron'.

In the file, add 
```
SYNAPSE_AUTH_TOKEN=xxx_your_token_here_xxx
```

### MacOS

TODO



## Test Your Installation

Restart R after adding your token. 
Then to test that your installation is working, load package and try to log in. 
If you set your authentication token successfully in the above, you should be greeted accordingly.

### Login

```{r}

library(nfportalutils)
syn_login()

```


