---
title: "Surveying public files in the portal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surveying public files in the portal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Intro

This quick makes use of some functions to survey files in the portal and their access. 

## Set up 

The usual setup:
```{r setup}
library(nfportalutils)
syn_login()
```

## Files downloadable for Synapse registered users

When talking about "public" files, this usually means files that are viewable and downloadable to Synapse users.
This group has id `273948`, so we use in the query below:

```{r query-1}
public_access <- summarize_file_access("DOWNLOAD", principal_id = 273948, "syn16858331")
public_access
```

Breakdown as absolute number and as proportions:
```{r query-1}
public_access[, .(n_downloadable = sum(N)), by = access][, .(access, n_downloadable, proportion = n_downloadable / sum(n_downloadable))]
```

## Some Nuances

While it would be nice to see the file access restrictions at different points in time, note that the underlying API only returns access control info at present.
A file may have inherited from a benefactor at an earlier point, but then becomes its own benefactor later (i.e. more granular access control),
so queries based on a past state will likely not work.
Don't try something like: 
```{r query-bad, eval=F}
public_access_q3_2022 <- summarize_file_access("DOWNLOAD", principal_id = 273948, "syn16858331.47")
```