---
title: "Validation workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{revalidation-workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This vignette demonstrates two approaches for programmatic metadata validation workflows using the utilities in this package. 
Validating metadata via the UI is covered elsewhere. 

First set up as usual:

```{r, eval = F}

library(nfportalutils)
syn_login(Sys.getenv("SYNAPSE_AUTH_TOKEN"))

```


## Option 1: Using Synapse validation service

The Synapse validation service will be the main validation service to use starting in 2026. 
Before validating with Synapse's native backend capabilities, make sure a JSON schema is "bound" to the folder. 
Here's our example folder with several files; two directly under the folder and one in a subfolder.

```{r}
my_dataset <- "syn51106460"
```

1. Bind the schema that we know to be the appropriate one: 

```{r, eval=F}

# Preferred: Bind explicit version
bind_schema(id = my_dataset, schema_id = "org.synapse.nf-generalmeasuredatatemplate-10.2.0")

# Use carefully: Bind the "latest" (last registered) version
# bind_schema(id = my_dataset, schema_id = "org.synapse.nf-generalmeasuredatatemplate")
```

2. Validate all items in the folder. 
A fileview that includes the folder in the scope is required.
The results are returned as the "raw" response; if you want a summary format, create a function to process the results as desired. 

```{r, eval=F}
validate_dataset_folder(my_dataset, fileview = "syn64364141")
```


## Option 2 (will be DEPRECATED in 2026): Using Schematic API service

Most validation will use schematic service until 2026. 
Note that the schematic API only works with dataset folders currently. 
An example dataset folder is given below:

```{r, eval=F}

my_dataset <- "syn25386362"

```

1. To validate metadata, a manifest must be reconstituted, i.e. we create a csv from the annotations currently on the files. 
Type `?manifest_generate` to read the docs.

You need to know the `data_type` to validate against.
The `data_type` is the same as the "Component" value in the schematic data model.
If feeling lucky, try `infer_data_type`.

```r

inferred <- infer_data_type(my_dataset)
inferred
#> $result
#> [1] "GenomicsAssayTemplate"
#> 
#> $notes
#> [1] ""
```

Once `data_type` is in hand, here's the general invocation:

```r

data_type <- inferred$result # Or set manually otherwise

manifest_generate(data_type, 
                  dataset_id = my_dataset,
                  schema_url = "https://raw.githubusercontent.com/nf-osi/nf-metadata-dictionary/main/NF.jsonld",
                  output_format = "google_sheet") # otherwise excel 
#> Manifest generated as Googlesheet(s)
#> [[1]]
#> [1] "https://docs.google.com/spreadsheets/d/11ymlnESzn7XhHS3vHzlRFsDPhafJJehlhuCL9HgICt8"
```

2. Go to the google_sheet and download as `.csv`. If Excel was chosen, open in some spreadsheet editor and resave file as `.csv`.
Then validate.

```r
manifest_validate(data_type = data_type, 
                  file_name = "GenomicsAssayTemplate - Sheet1.csv")
#> $errors
#> $errors[[1]]
#> $errors[[1]][[1]]
#> [1] "2"
#> 
#> $errors[[1]][[2]]
#> [1] "assay"
#> 
#> $errors[[1]][[3]]
#> [1] "'' is not one of ['NIH Toolbox', 'STR profile', 'traction force microscopy', 'massively parallel reporter assay', 'gait measurement', 'conventional MRI', 'functional MRI', 'immunoassay', 'contextual conditioning behavior assay', 'genotyping', 'DNA optical mapping', 'NOMe-seq', 'Social Responsiveness Scale', 'targeted exome sequencing', '2D AlamarBlue fluorescence', 'TMT quantitation', 'liquid chromatography-electrochemical detection', 'whole genome sequencing', 'Riccardi and Ablon scales', 'cell"
#> 
#> $errors[[1]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[2]]
#> $errors[[2]][[1]]
#> [1] "2"
#> 
#> $errors[[2]][[2]]
#> [1] "specimenID"
#> 
#> $errors[[2]][[3]]
#> [1] "'' is too short"
#> 
#> $errors[[2]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[3]]
#> $errors[[3]][[1]]
#> [1] "2"
#> 
#> $errors[[3]][[2]]
#> [1] "libraryStrand"
#> 
#> $errors[[3]][[3]]
#> [1] "'' is not one of ['FirstStranded', 'Unstranded', 'Not Applicable', 'SecondStranded']"
#> 
#> $errors[[3]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[4]]
#> $errors[[4]][[1]]
#> [1] "2"
#> 
#> $errors[[4]][[2]]
#> [1] "tumorType"
#> 
#> $errors[[4]][[3]]
#> [1] "'' is not one of ['Anaplastic Ganglioglioma', 'Anaplastic Astrocytoma', 'Nodular Neurofibroma', 'Meningioma', 'Fibrosarcoma', 'Localized Neurofibroma', 'Glioblastoma', 'Malignant Peripheral Nerve Sheath Tumor', 'Anaplastic Pleomorphic Xanthoastrocytoma', 'Atypical Neurofibroma', 'tumor', 'Colorectal Adenocarcinoma', 'Recurrent MPNST', 'Pilocytic Astrocytoma', 'Ganglioglioma', 'Optic Pathway Glioma', 'Neurofibroma', 'Necrotic Neoplasm', 'Glioma', 'Teratoma', 'Cutaneous Neurofibroma', 'Fibromatosi"
#> 
#> $errors[[4]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[5]]
#> $errors[[5]][[1]]
#> [1] "2"
#> 
#> $errors[[5]][[2]]
#> [1] "libraryPreparationMethod"
#> 
#> $errors[[5]][[3]]
#> [1] "'' is not one of ['CEL-seq', 'NEBNext mRNA Library Prep Reagent Set for Illumina', '10x', 'GTAC@WUSTL in-house prep', 'KAPA mRNA HyperPrep Kit', 'TruSeq', 'unknown', 'KAPA HyperPrep Kit PCR-free', 'Illumina TruSeq DNA Nano', 'TruSeq standard total RNA library kit', 'QuantSeq FWD V2 with UDI', 'Drop-Seq', 'KAPA RNA HyperPrep Kit with RiboErase (HMR)', 'Smart-seq4', 'IDT xGen Exome Research Panel', 'Smart-seq2', 'Omni-ATAC']"
#> 
#> $errors[[5]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[6]]
#> $errors[[6]][[1]]
#> [1] "2"
#> 
#> $errors[[6]][[2]]
#> [1] "individualID"
#> 
#> $errors[[6]][[3]]
#> [1] "'' is too short"
#> 
#> $errors[[6]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[7]]
#> $errors[[7]][[1]]
#> [1] "2"
#> 
#> $errors[[7]][[2]]
#> [1] "platform"
#> 
#> $errors[[7]][[3]]
#> [1] "'' is not one of ['Illumina Genome Analyzer IIx', 'Illumina HiSeq X', 'Perlegen 300Karray', 'Vevo 3100 Imaging System', 'Illumina MouseWG-6 v2.0 expression beadchip', 'Illumina Infinium MethylationEPIC BeadChip v2.0 (935k)', 'Vectra H1 3D Imaging System', 'Nanostring Counter', 'Illumina Infinium MethylationEPIC BeadChip v1.0 (850k)', 'Illumina HumanOmniExpress-24 v1.0 BeadChip', 'Illumina HumanOmni1-Quadv1.0', 'LifeViz Micro System', 'LI-COR Odyssey CLx', 'Illumina HumanMethylation450', 'Illumin"
#> 
#> $errors[[7]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[8]]
#> $errors[[8]][[1]]
#> [1] "2"
#> 
#> $errors[[8]][[2]]
#> [1] "specimenPreparationMethod"
#> 
#> $errors[[8]][[3]]
#> [1] "'' is not one of ['FFPE', 'OCT', 'RNAlater', 'Viably frozen', 'Fresh collected', 'Cryopreserved', 'formalin-fixed', 'Flash frozen', 'ethanol']"
#> 
#> $errors[[8]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[9]]
#> $errors[[9]][[1]]
#> [1] "2"
#> 
#> $errors[[9]][[2]]
#> [1] "species"
#> 
#> $errors[[9]][[3]]
#> [1] "'' is not one of ['Rattus norvegicus', 'Gallus gallus', 'Danio rerio', 'Sus scrofa', 'Drosophila melanogaster', 'Oryctolagus cuniculus', 'Pan troglodytes', 'Rhesus macaque', 'Mus musculus (humanized)', 'Homo sapiens', 'Mus musculus']"
#> 
#> $errors[[9]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[10]]
#> $errors[[10]][[1]]
#> [1] "3"
#> 
#> $errors[[10]][[2]]
#> [1] "assay"
#> 
#> $errors[[10]][[3]]
#> [1] "'' is not one of ['NIH Toolbox', 'STR profile', 'traction force microscopy', 'massively parallel reporter assay', 'gait measurement', 'conventional MRI', 'functional MRI', 'immunoassay', 'contextual conditioning behavior assay', 'genotyping', 'DNA optical mapping', 'NOMe-seq', 'Social Responsiveness Scale', 'targeted exome sequencing', '2D AlamarBlue fluorescence', 'TMT quantitation', 'liquid chromatography-electrochemical detection', 'whole genome sequencing', 'Riccardi and Ablon scales', 'cell"
#> 
#> $errors[[10]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[11]]
#> $errors[[11]][[1]]
#> [1] "3"
#> 
#> $errors[[11]][[2]]
#> [1] "specimenID"
#> 
#> $errors[[11]][[3]]
#> [1] "'' is too short"
#> 
#> $errors[[11]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[12]]
#> $errors[[12]][[1]]
#> [1] "3"
#> 
#> $errors[[12]][[2]]
#> [1] "libraryStrand"
#> 
#> $errors[[12]][[3]]
#> [1] "'' is not one of ['FirstStranded', 'Unstranded', 'Not Applicable', 'SecondStranded']"
#> 
#> $errors[[12]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[13]]
#> $errors[[13]][[1]]
#> [1] "3"
#> 
#> $errors[[13]][[2]]
#> [1] "tumorType"
#> 
#> $errors[[13]][[3]]
#> [1] "'' is not one of ['Anaplastic Ganglioglioma', 'Anaplastic Astrocytoma', 'Nodular Neurofibroma', 'Meningioma', 'Fibrosarcoma', 'Localized Neurofibroma', 'Glioblastoma', 'Malignant Peripheral Nerve Sheath Tumor', 'Anaplastic Pleomorphic Xanthoastrocytoma', 'Atypical Neurofibroma', 'tumor', 'Colorectal Adenocarcinoma', 'Recurrent MPNST', 'Pilocytic Astrocytoma', 'Ganglioglioma', 'Optic Pathway Glioma', 'Neurofibroma', 'Necrotic Neoplasm', 'Glioma', 'Teratoma', 'Cutaneous Neurofibroma', 'Fibromatosi"
#> 
#> $errors[[13]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[14]]
#> $errors[[14]][[1]]
#> [1] "3"
#> 
#> $errors[[14]][[2]]
#> [1] "libraryPreparationMethod"
#> 
#> $errors[[14]][[3]]
#> [1] "'' is not one of ['CEL-seq', 'NEBNext mRNA Library Prep Reagent Set for Illumina', '10x', 'GTAC@WUSTL in-house prep', 'KAPA mRNA HyperPrep Kit', 'TruSeq', 'unknown', 'KAPA HyperPrep Kit PCR-free', 'Illumina TruSeq DNA Nano', 'TruSeq standard total RNA library kit', 'QuantSeq FWD V2 with UDI', 'Drop-Seq', 'KAPA RNA HyperPrep Kit with RiboErase (HMR)', 'Smart-seq4', 'IDT xGen Exome Research Panel', 'Smart-seq2', 'Omni-ATAC']"
#> 
#> $errors[[14]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[15]]
#> $errors[[15]][[1]]
#> [1] "3"
#> 
#> $errors[[15]][[2]]
#> [1] "individualID"
#> 
#> $errors[[15]][[3]]
#> [1] "'' is too short"
#> 
#> $errors[[15]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[16]]
#> $errors[[16]][[1]]
#> [1] "3"
#> 
#> $errors[[16]][[2]]
#> [1] "platform"
#> 
#> $errors[[16]][[3]]
#> [1] "'' is not one of ['Illumina Genome Analyzer IIx', 'Illumina HiSeq X', 'Perlegen 300Karray', 'Vevo 3100 Imaging System', 'Illumina MouseWG-6 v2.0 expression beadchip', 'Illumina Infinium MethylationEPIC BeadChip v2.0 (935k)', 'Vectra H1 3D Imaging System', 'Nanostring Counter', 'Illumina Infinium MethylationEPIC BeadChip v1.0 (850k)', 'Illumina HumanOmniExpress-24 v1.0 BeadChip', 'Illumina HumanOmni1-Quadv1.0', 'LifeViz Micro System', 'LI-COR Odyssey CLx', 'Illumina HumanMethylation450', 'Illumin"
#> 
#> $errors[[16]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[17]]
#> $errors[[17]][[1]]
#> [1] "3"
#> 
#> $errors[[17]][[2]]
#> [1] "specimenPreparationMethod"
#> 
#> $errors[[17]][[3]]
#> [1] "'' is not one of ['FFPE', 'OCT', 'RNAlater', 'Viably frozen', 'Fresh collected', 'Cryopreserved', 'formalin-fixed', 'Flash frozen', 'ethanol']"
#> 
#> $errors[[17]][[4]]
#> [1] ""
#> 
#> 
#> $errors[[18]]
#> $errors[[18]][[1]]
#> [1] "3"
#> 
#> $errors[[18]][[2]]
#> [1] "species"
#> 
#> $errors[[18]][[3]]
#> [1] "'' is not one of ['Rattus norvegicus', 'Gallus gallus', 'Danio rerio', 'Sus scrofa', 'Drosophila melanogaster', 'Oryctolagus cuniculus', 'Pan troglodytes', 'Rhesus macaque', 'Mus musculus (humanized)', 'Homo sapiens', 'Mus musculus']"
#> 
#> $errors[[18]][[4]]
#> [1] ""
#> 
#> 
#> 
#> $warnings
#> $warnings[[1]]
#> $warnings[[1]][[1]]
#> [1] "2"
#> 
#> $warnings[[1]][[2]]
#> [1] "age"
#> 
#> $warnings[[1]][[3]]
#> [1] "On row 2 the attribute age does not contain the proper value type num."
#> 
#> $warnings[[1]][[4]]
#> [1] ""
#> 
#> 
#> $warnings[[2]]
#> $warnings[[2]][[1]]
#> [1] "3"
#> 
#> $warnings[[2]][[2]]
#> [1] "age"
#> 
#> $warnings[[2]][[3]]
#> [1] "On row 3 the attribute age does not contain the proper value type num."
#> 
#> $warnings[[2]][[4]]
#> [1] ""
#> 
#> 
#> $warnings[[3]]
#> $warnings[[3]][[1]]
#> $warnings[[3]][[1]][[1]]
#> [1] "2"
#> 
#> $warnings[[3]][[1]][[2]]
#> [1] "3"
#> 
#> 
#> $warnings[[3]][[2]]
#> [1] "readPair"
#> 
#> $warnings[[3]][[3]]
#> [1] "readPair values in rows ['2', '3'] are out of the specified range."
#> 
#> $warnings[[3]][[4]]
#> $warnings[[3]][[4]][[1]]
#> [1] ""
#> 
#> 
#> 
#> $warnings[[4]]
#> $warnings[[4]][[1]]
#> [1] "2"
#> 
#> $warnings[[4]][[2]]
#> [1] "readLength"
#> 
#> $warnings[[4]][[3]]
#> [1] "On row 2 the attribute readLength does not contain the proper value type int."
#> 
#> $warnings[[4]][[4]]
#> [1] ""
#> 
#> 
#> $warnings[[5]]
#> $warnings[[5]][[1]]
#> [1] "3"
#> 
#> $warnings[[5]][[2]]
#> [1] "readLength"
#> 
#> $warnings[[5]][[3]]
#> [1] "On row 3 the attribute readLength does not contain the proper value type int."
#> 
#> $warnings[[5]][[4]]
#> [1] ""
#> 
#> 
#> $warnings[[6]]
#> $warnings[[6]][[1]]
#> [1] "2"
#> 
#> $warnings[[6]][[2]]
#> [1] "readDepth"
#> 
#> $warnings[[6]][[3]]
#> [1] "On row 2 the attribute readDepth does not contain the proper value type int."
#> 
#> $warnings[[6]][[4]]
#> [1] ""
#> 
#> 
#> $warnings[[7]]
#> $warnings[[7]][[1]]
#> [1] "3"
#> 
#> $warnings[[7]][[2]]
#> [1] "readDepth"
#> 
#> $warnings[[7]][[3]]
#> [1] "On row 3 the attribute readDepth does not contain the proper value type int."
#> 
#> $warnings[[7]][[4]]
#> [1] ""
#> 
#> 
#> $warnings[[8]]
#> $warnings[[8]][[1]]
#> [1] "2"
#> 
#> $warnings[[8]][[2]]
#> [1] "experimentalTimepoint"
#> 
#> $warnings[[8]][[3]]
#> [1] "On row 2 the attribute experimentalTimepoint does not contain the proper value type num."
#> 
#> $warnings[[8]][[4]]
#> [1] ""
#> 
#> 
#> $warnings[[9]]
#> $warnings[[9]][[1]]
#> [1] "3"
#> 
#> $warnings[[9]][[2]]
#> [1] "experimentalTimepoint"
#> 
#> $warnings[[9]][[3]]
#> [1] "On row 3 the attribute experimentalTimepoint does not contain the proper value type num."
#> 
#> $warnings[[9]][[4]]
#> [1] ""
```

3. Make corrections in the `.csv` according to validation laundry list.

4. Submit corrected manifest via DCA (or via `annotate_with_manifest`).

